Gammapy: A Python package for gamma-ray astronomy

### 2023-03-15 src/ms-review.tex
Address more review comments

%% CN: what about mentioning the joint Crab? or is it mentioned later?
With the public availability of the \gadf format specifications and the
\gammapy package, some experiments started to make limited subsets of
their \gammaray data publicly available for testing and validating
\gammapy. For example, the \hess collaboration released a limited test
dataset (about 50 hours of observations taken between 2004 and 2008)
based on the \gadf DL3 format \citep{HESS_DR1}. This data release served
as a basis for validation of open analysis tools, including \gammapy 
\cite[see e.g.][]{Mohrmann2019}. The \hawc collaboration also released
a limited test dataset of the Crab Nebula, which was used to validate
the \gammapy package in \cite{Olivera2022}.
The increased availability of public data that followed the definition of a common data format, and the development of \gammapy as a community-driven open software, led the way toward a more open science in the very-high-energy \gray Astronomy domain. In future CTAO will be an open observatory committed to follow the FAIR (Findable, Accessible, Interoperable and Reusable) principles \citep{FAIR16, FAIR22} that define the key requirements for open science.

@@

\subsection{gammapy.data}
\label{ssec:gammapy-data}
The \code{gammapy.data} sub-package implements the functionality to select,
read, and represent DL3 \gammaray data in memory. It provides the main user
interface to access the lowest data level. \gammapy currently only
supports data that is compliant with \code{v0.2} and \code{v0.3} of the \gadf data format.
DL3 data are typically bundled into individual observations, corresponding to stable periods of data acquisition. For IACT data analysis,
for which the GADF data model and Gammapy were initially conceived,
these are usually $20 - 30\,{\rm min}$ long.
Each observation is assigned a unique integer ID for reference.

@@

\section{Applications}
\label{sec:applications}
%% CN: I think we never explained what gammapy-data is and we should have at this point 
%% (it is used without mention also in the TSEstimator example).
\gammapy is currently used for a variety of analyses by different IACT
experiments and has already been employed in more than 60 scientific publications as of XX/03/2023
\footnote{\href{https://ui.adsabs.harvard.edu/search/q=(\%20(citations(doi\%3A\%2210.1051\%2F0004-6361\%2F201834938\%22)\%20OR\%20citations(bibcode\%3A2017ICRC...35..766D))\%20AND\%20year\%3A2014-2023)&sort=date\%20desc\%2C\%20bibcode\%20desc&p_=0}{List on ADS}}.
In this section, we illustrate the capabilities of \gammapy by performing some standard
analysis cases commonly considered in \gammaray astronomy.
Beside reproducing standard methodologies, we illustrate the unique data combination
capabilites of Gammapy by presenting a multi-instrument analysis to date not possible within any
of the current instrument private software frameworks.
The examples shown are based on the data accessible in the \code{gammapy-data} repository,
and limited by the availability of public data.
We remark that, as long as the data are compliant with the GADF specifications (or its future evolutions),
and hence with Gammapy's data structures, there is no limitation on performing
analyses of data from a given instrument.

@@

\begin{figure*}
	\centering
	\includegraphics[width=1.\textwidth]{figures/cta_galactic_center.pdf}
	\caption{
		Example of a one dimensional spectral analysis of the Galactic Center for three simulated CTA
		observations for the 1DC dataset. The left image shows the maps of counts with the signal
		region in white and background regions overlaid in different colors. The right image
		shows the resulting spectral points and their corresponding log-likelihood
		profiles.}
	\label{fig:cta_galactic_center}
\end{figure*}
%% CN: expand 1DC acronym and add a reference, here or in the section text

@@

\subsection{Temporal Analysis}
\label{ssec:temporal-analysis}
A common use case in many astrophysical scenarios is to study the temporal
variability of a source. The most basic way to do this is to construct a
light curve, i.e., the flux of a source in each given time bin. In \gammapy, this
is done by using the \code{LightCurveEstimator} that fits the normalisation of a
source in each time (and optionally energy) band per observation, keeping constant
other parameters.
For custom time binning, an observation needs to be split into finer time bins using
the \code{Observation.select\_time} method. Figure~\ref{fig:hess_lightcurve_pks}
shows the light curve of the blazar PKS~2155-304 in different energy bands as
observed by the \hess telescope during an exceptional flare on the night of
July 29 - 30, 2006~\cite{2009A&A...502..749A}. The data are publicly available 
as a part of the HESS-DL3-DR1~\cite{HESS_DR1}. Each observation is first split into 10 min smaller
observations, and spectra extracted for each of these within a 0.11\textdegree~radius
around the source. A \code{PowerLawSpectralModel} is fit to all the datasets, leading
to a reconstructed index of $3.54 \pm 0.02$. With this adjusted spectral model
the \code{LightCurveEstimator} runs directly for two energy bands,
\SI{0.5}{TeV}~to~{1.5}{TeV} and \SI{1.5}{TeV}~to~{20}{TeV} respectively.
%
\begin{figure*}[t]
    \sidecaption
	\includegraphics[width=0.6666\textwidth]{figures/hess_lightcurve_pks.pdf}
	\caption{
        Binned light curves in two different energy bands for the source
        PKS~2155-304 in two energy bands (\SI{0.5}{TeV}~to~{1.5}{TeV} and \SI{1.5}{TeV}~to~{20}{TeV})
        as observed by the \hess telescopes in 2006. The coloured markers
        show the flux points in the different energy bands. The horizontal
        error illustrates the width of the time bin of 10~min. The vertical
        error bars show the associated asymmetrical flux errors. The marker
        is set to the center of the time bin.
    }
    \label{fig:hess_lightcurve_pks}
\end{figure*}
%
The obtained flux points can be analytically modelled using the available or
user-implemented temporal models. Alternatively, instead of  extracting a
light curve, it is also possible to directly fit temporal models to the reduced
datasets. By associating an appropriate \code{SkyModel}, consisting of both temporal
and spectral components, or using custom temporal models with spectroscopic
variability, to each dataset, a joint fit across the datasets will directly
return the best fit temporal and spectral parameters.

@@

\subsection{Broadband SED Modelling}
\label{ssec:broadband-sed-modeling}
By combining \gammapy with astrophysical modelling codes, users can also fit
astrophysical spectral models to \gammaray data. In \gammaray
astronomy one typically observes two radiation production
mechanisms, the so-called hadronic and leptonic scenarios.
There are several Python packages that are able to model
the \gammaray emission, given a physical scenario. Among those
packages are Agnpy~\citep{agnpy}, Naima~\citep{naima}, Jetset~\citep{jetset}
and Gamera~\citep{gamera}.
Typically those emission models predict broadband emission from
radio, up to the very high-energy \gammaray range.
By relying on the multiple dataset types in \gammapy those
data can be combined to constrain such a broadband emission model.
\gammapy provides a built-in \code{NaimaSpectralModel} that allows
users to wrap a given astrophysical emission model from the
Naima package and fit it directly to \gammaray data.

As an example of this application, we use the same multi-instrument
dataset described in the previous section and we fit it with an inverse
Compton model computed with Naima and wrapped in the \gammapy models
through the \code{NaimaSpectralModel} class. We describe the gamma-ray emission 
with an inverse Compton scenario, considering a log-parabolic
electron distribution that scatters: the synchrotron radiation produced by
the very same electrons; near and far infrared photon fields and the cosmic
microwave background (CMB). We adopt the prescription on the target
photon fields provided in the documentation of the \textit{Naima}
package\footnote{\url{https://naima.readthedocs.io/en/stable/examples.html\#crab-nebula-ssc-model}}.
The best-fit inverse Compton spectrum is represented with a red dashed line in
Figure~\ref{fig:multi_instrument_analysis}.

@@

	We would like to thank the \texttt{Numpy}, \texttt{Scipy}, \texttt{IPython} and
	\texttt{Matplotlib} communities for providing their packages which are
	invaluable to the development of Gammapy. We thank the \github team for
	providing us with an excellent free development platform. We also are grateful
	to Read the Docs (\ReadthedocsUrl), and Travis (\TravisUrl) for providing free
	documentation hosting and testing respectively. Α special acknowledgment has to be given
	to our first Lead Developer of \gammapy, Christoph Deil. Finally, we would like to thank
	all the \gammapy users that have provided feedback and submitted bug reports.

>>>>

%% CN: what about mentioning the joint Crab? or is it mentioned later?
With the public availability of the \gadf format specifications and the
\gammapy package, some experiments started to make limited subsets of
their \gammaray data publicly available for testing and validating
\gammapy. For example, the \hess collaboration released a limited test
dataset (about 50 hours of observations taken between 2004 and 2008)
based on the \gadf DL3 format \citep{HESS_DR1}. This data release served
as a basis for validation of open analysis tools, including \gammapy 
\cite[see e.g.][]{Mohrmann2019}. The \hawc collaboration also released
a limited test dataset of the Crab Nebula, which was used to validate
the \gammapy package in \cite{Olivera2022}.
The increased availability of public data that followed the definition of
a common data format, and the development of \gammapy as a community-driven open software,
led the way toward a more open science in the very-high-energy \gammaray Astronomy domain.
In future CTAO will be an open observatory committed to follow the FAIR
(Findable, Accessible, Interoperable and Reusable) principles \citep{FAIR16, FAIR22} 
that define the key requirements for open science.

@@

\subsection{gammapy.data}
\label{ssec:gammapy-data}
The \code{gammapy.data} sub-package implements the functionality to select,
read, and represent DL3 \gammaray data in memory. It provides the main user
interface to access the lowest data level. \gammapy currently only
supports data that is compliant with \code{v0.2} and \code{v0.3} of the \gadf data format.
DL3 data are typically bundled into individual observations, corresponding to stable periods of data acquisition. For IACT data analysis,
for which the GADF data model and \gammapy were initially conceived,
these are usually $20 - 30\,{\rm min}$ long.
Each observation is assigned a unique integer ID for reference.

@@

\section{Applications}
\label{sec:applications}
%% CN: I think we never explained what gammapy-data is and we should have at this point 
%% (it is used without mention also in the TSEstimator example).
\gammapy is currently used for a variety of analyses by different IACT
experiments and has already been employed in more than 60 scientific publications as of XX/03/2023
\footnote{\href{https://ui.adsabs.harvard.edu/search/q=(\%20(citations(doi\%3A\%2210.1051\%2F0004-6361\%2F201834938\%22)\%20OR\%20citations(bibcode\%3A2017ICRC...35..766D))\%20AND\%20year\%3A2014-2023)&sort=date\%20desc\%2C\%20bibcode\%20desc&p_=0}{List on ADS}}.
In this section, we illustrate the capabilities of \gammapy by performing some standard
analysis cases commonly considered in \gammaray astronomy.
Beside reproducing standard methodologies, we illustrate the unique data combination
capabilites of \gammapy by presenting a multi-instrument analysis, which is not possible within any
of the current instrument private software frameworks.
The examples shown are based on the data accessible in the \code{gammapy-data} repository,
and limited by the availability of public data.
We remark that, as long as the data are compliant with the GADF specifications (or its future evolutions),
and hence with Gammapy's data structures, there is no limitation on performing
analyses of data from a given instrument.

@@

\begin{figure*}
	\centering
	\includegraphics[width=1.\textwidth]{figures/cta_galactic_center.pdf}
	\caption{
		Example of a one dimensional spectral analysis of the Galactic Center for three simulated CTA
		observations from the 1DC dataset. The left image shows the maps of counts with the signal
		region in white and the reflected background regions for the three different observations
		overlaid in different colors. The right image shows the resulting spectral flux points and
		their corresponding log-likelihood profiles. The flux points are shown in orange, with the
		horizontal bar illustrating the width of the energy bin and the vertical bar the $1~\sigma$
		error. The log-likelihood profiles for each enetgy bin are shown in the background. The colormap
		illustrates the difference of the log-likelihood to the log-likelihood of the best fit value.
		}
	\label{fig:cta_galactic_center}
\end{figure*}
%% CN: expand 1DC acronym and add a reference, here or in the section text

@@

\subsection{Temporal Analysis}
\label{ssec:temporal-analysis}
A common use case in many astrophysical scenarios is to study the temporal
variability of a source. The most basic way to do this is to construct a
light curve, i.e., the flux of a source in each given time bin. In \gammapy, this
is done by using the \code{LightCurveEstimator} that fits the normalisation of a
source in each time (and optionally energy) band per observation, keeping constant
other parameters.
For custom time binning, an observation needs to be split into finer time bins using
the \code{Observation.select\_time} method. Figure~\ref{fig:hess_lightcurve_pks}
shows the light curve of the blazar PKS~2155-304 in different energy bands as
observed by the \hess telescope during an exceptional flare on the night of
July 29 - 30, 2006~\cite{2009A&A...502..749A}. The data are publicly available 
as a part of the HESS-DL3-DR1~\cite{HESS_DR1}. Each observation is first split into 10 min smaller
observations, and spectra extracted for each of these within a 0.11\textdegree~radius
around the source. A \code{PowerLawSpectralModel} is fit to all the datasets, leading
to a reconstructed index of $3.54 \pm 0.02$. With this adjusted spectral model
the \code{LightCurveEstimator} runs directly for two energy bands,
\SI{0.5}{TeV}~to~{1.5}{TeV} and \SI{1.5}{TeV}~to~{20}{TeV} respectively.
%
\begin{figure*}[t]
    \sidecaption
	\includegraphics[width=0.6666\textwidth]{figures/hess_lightcurve_pks.pdf}
	\caption{
        Binned PKS~2155-304 light curve in two different energy bands
        as observed by the \hess telescopes in 2006. The coloured markers
        show the flux points in the different energy bands:
		the range from (\SI{0.5}{TeV}~to~{1.5}{TeV} is shown in blue, while
		the range from \SI{1.5}{TeV}~to~{20}{TeV}) is shown in orange.
		The horizontal error illustrates the width of the time bin of 10~min. The vertical
        error bars show the associated asymmetrical flux errors. The marker
        is set to the center of the time bin.
    }
    \label{fig:hess_lightcurve_pks}
\end{figure*}
%
The obtained flux points can be analytically modelled using the available or
user-implemented temporal models. Alternatively, instead of  extracting a
light curve, it is also possible to directly fit temporal models to the reduced
datasets. By associating an appropriate \code{SkyModel}, consisting of both temporal
and spectral components, or using custom temporal models with spectroscopic
variability, to each dataset, a joint fit across the datasets will directly
return the best fit temporal and spectral parameters.

@@

\subsection{Broadband SED Modelling}
\label{ssec:broadband-sed-modeling}
By combining \gammapy with astrophysical modelling codes, users can also fit
astrophysical spectral models to \gammaray data. 
There are several Python packages that are able to model
the \gammaray emission, given a physical scenario. Among those
packages are Agnpy~\citep{agnpy}, Naima~\citep{naima}, Jetset~\citep{jetset}
and Gamera~\citep{gamera}.
Typically those emission models predict broadband emission from
radio, up to very high-energy \gammarays.
By relying on the multiple dataset types in \gammapy those
data can be combined to constrain such a broadband emission model.
\gammapy provides a built-in \code{NaimaSpectralModel} that allows
users to wrap a given astrophysical emission model from the
Naima package and fit it directly to \gammaray data.

As an example application, we use the same multi-instrument
dataset of the Crab Nebula, described in the previous section,
and we apply an inverse
Compton model computed with Naima and wrapped in the \gammapy models
through the \code{NaimaSpectralModel} class. We describe the gamma-ray emission 
with an inverse Compton scenario, considering a log-parabolic
electron distribution that scatters photons from:
%
\begin{itemize}
	\item the synchrotron radiation produced by
	the very same electrons
	\item near and far infrared photon fields
	\item and the cosmic microwave background (CMB)
\end{itemize}
%
We adopt the prescription on the target photon fields provided in the documentation of the \textit{Naima}
package\footnote{\url{https://naima.readthedocs.io/en/stable/examples.html\#crab-nebula-ssc-model}}.
The best-fit inverse Compton spectrum is represented with a red dashed line in
Figure~\ref{fig:multi_instrument_analysis}.

@@

	We would like to thank the \texttt{Numpy}, \texttt{Scipy}, \texttt{IPython} and
	\texttt{Matplotlib} communities for providing their packages which are
	invaluable to the development of \gammapy. We thank the \github team for
	providing us with an excellent free development platform. We also are grateful
	to Read the Docs (\ReadthedocsUrl), and Travis (\TravisUrl) for providing free
	documentation hosting and testing respectively. Α special acknowledgment has to be given
	to our first Lead Developer of \gammapy, Christoph Deil. Finally, we would like to thank
	all the \gammapy users that have provided feedback and submitted bug reports.
